Measurement Bias from Address Aliasing

### 2014-05-21 paper.tex
Prune references

% This is basically former background section
Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~%\cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Intel:2013:Volume3B}.
A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{Intel:2012:OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item[LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS]
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{Intel:2012:OptimizationManual}
\end{description}

@@

On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Intel:2013:Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
Our assumption is that this is a by-product of stalls generated from aliasing, causing micro-operations to be reissued.

>>>>

% This is basically former background section
Bias in performance analysis has been shown to be commonplace and significant in real applications, while being unpredictable and difficult to deal with~% todo: \cite{Mytkowicz:2008:OE&MB, Mytkowicz:2009:WrongData, Mytkowicz:2008:Easy}.
Previous work looks at environmental factors such as the size of Unix environment variables, which can affect memory by offsetting the initial position of stack.
Because of the complexity of modern hardware, it is often impossible to predict exactly how things such as environment variables ultimately affect program execution.
Without an accurate understanding of the underlying memory interactions, the best one can do to accurately assess performance is generally to average over a large sample set of execution contexts.
Previous work focuses mostly of how to account for bias by proper sampling, using randomization of execution contexts and statistical methods. % Cite is basically the same as above, all three are about the same -- difficult to split.

@@

Performance counters are supported in the Linux kernel, and we access them via a tool called \emph{perf}. 
We use the \texttt{perf-stat} command in all experiments, which accepts raw event codes listed in the Intel reference manual \cite{Volume3B}.
A small Python script is used to collect an exhaustive set of all available counters, which amounts to about 200 on our architecture.
The same script simulates changes to environment size, which is provoked by setting a dummy environment variable to $n$ number of zero characters, starting from a minimal environment.\footnote{Because perf-stat itself adds a few variables, the environment will never be completely empty.}
Interesting events are identified by computing linear correlation to cycle count, measuring all counters over a series of execution contexts.
Cache related metrics are monitored in order to rule out cache as the underlying cause of bias, such as hit rates of micro-ops for each level of cache~\cite{OptimizationManual}.
Results are averaged over multiple runs to reduce potential random error, supplying the \texttt{-r} option to perf-stat. % Not median, good question. man perf-stat says average + std.dev.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, and causing the load to be reissued.
The number of times this happens can be counted by the following performance counter:
\begin{description}
  \item[LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS]
  Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.~\cite{OptimizationManual} % todo: where
\end{description}

@@

On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from about 270,000 to about 135,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports seem to have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change, but for some reason the micro-ops stays in the execution pipeline for a shorter time.
Our assumption is that this is a by-product of stalls generated from aliasing, causing micro-operations to be reissued.
